---
title: "Final Submission"
output: pdf_document
date: "2024-07-29"
---
```{r}
setwd("/Users/yanchuwu/Downloads") # set wd

#read csv
genes <- read.csv("QBS103_GSE157103_genes.csv",row.names = 1 ) # row name are coming from 1st colum of csv
new_genes<- t(genes)
#transform the rows to column
```

```{r}
setwd("/Users/yanchuwu/Downloads") # set wd

#read csv

series_matrix <- read.csv("QBS103_GSE157103_series_matrix.csv")



# merge data together in to one data set file
merged_data<- merge(new_genes, series_matrix, by.x = "row.names", by.y = "participant_id", all = TRUE)

```

# categorical data
sex
icu_status
#continues
age
#gene
A1BG

#gene expression

```{r}
library(dplyr)
library(ggplot2) # required library

```

#histogram
```{r}
# I picked the gene A1BG as my x 

library(ggplot2)
my_histogram<-ggplot(merged_data,aes(x = A1BG)) + 
  geom_histogram(binwidth = 0.1, fill = "blue", alpha = 0.7) +
  labs(title = "Gene(A1BG) Expression Histogram", x = "Gene(A1BG)", y = "Frequency")

my_histogram
  
```
#save my histogram
```{r}
ggsave("my_histogram.pdf", plot = my_histogram, width = 8, height = 6)
```



#scatterplot
```{r}
# age is my continuous covariate 
library(ggplot2)
merged_data$age <- as.numeric(merged_data$age)
my_scatterplot<-ggplot(merged_data,aes(x = age ,y = A1BG ,color = age)) + # The + is how we communicate to ggplot2 that we aren't done adding elements to the plot
  # Add points to our scatter plot
  
  geom_point()+
  
  labs(x = 'Age(year)',y = 'Gene(A1BG)',color = 'Age(year)', title = 'Age vs Gene(A1BG) scatterplot' ) 

my_scatterplot

#age and gene expression does not have strong correlation, no very obvious trend 
```
```{r}
ggsave("my_scatterplot.pdf", plot = my_scatterplot, width = 8, height = 6)
```


#asking for to visualize , gene as y axis, x axis has 2 categorical, 1 is color, and i 

```{r}
#gene expression separated by both categorical covariates(sex as x, fill as icu_status)
#boxplot
my_boxplot<-ggplot(merged_data,aes(x = sex,y = A1BG, fill = icu_status)) +
  # Add box plot
  geom_boxplot() +
  # Change labels
  labs(x = 'age',y = 'sex',color = 'icustatus', title = 'age vs sex boxplot' ) 
# white line is a median, bos size shows the variability
my_boxplot
```
save m box plot
```{r}
ggsave("my_boxplot.pdf", plot = my_boxplot, width = 8, height = 6)
```



#Generate a heatmap (5 pts)
Heatmap should include at least 10 genes
Include tracking bars for the 2 categorical covariates in your boxplot
Heatmaps should include clustered rows and columns


```{r}
library(pheatmap)
library(dbplyr)
# Subset the genes
gene_subset<-c("A1BG","A2M","A4GNT","AAAS","AACS","AAGAB","AAK1","AAMP","AANAT","AAR2")

gene_data_subset <- merged_data %>% select(gene_subset )

gene_data <- as.data.frame(t(gene_data_subset))



gene_data<-gene_data[-6]
print(gene_data)

annotation_data <- as.data.frame(merged_data %>% select("sex","disease_status"))  

print(annotation_data)

annotation_data <-annotation_data[-7,]


row.names(annotation_data)<-colnames(gene_data)


# make the annotation data matches the cleaned data



 annotationColors <- list(sex = c('female' = 'purple',
                                     'male' = 'blue',                                  'unknown' = 'red'),
                        icu_status = c('yes' = 'red',
                                        'no' = 'pink'))

# heatmap
my_pheatmap<-pheatmap(gene_data,
         clustering_rows= T,
         clustering_cols = T,
         show_colnames = FALSE,
         annotation_col = annotation_data)
         #annotation_colors = annotationColors)

my_pheatmap

ggsave("my_pheatmap.pdf", plot = my_pheatmap, width = 8, height = 6)

```

#summarry table
Stratifying by one of your categorical variables
Tables should report n (%) for categorical variables
Tables should report mean (sd) or median [IQR] for continuous variables

```{r}
merged_data<- merge(new_genes, series_matrix, by.x = "row.names", by.y = "participant_id", all = TRUE)

```
#categorical variable:sex,icu_status

Tables should report n (%) for categorical variables


```{r}
#data preparation
library(dplyr)
library(table1)

merged_data2 <- na.omit(merged_data)

merged_data2_filter <- merged_data2 %>% filter(!is.na(sex) & !is.na(icu_status) &! is.na(disease_status) &!is.na(age) &! is.na(ventilator.free_days) &! is.na(hospital.free_days_post_45_day_followup) &age==as.integer(age))

merged_data2_filtered <- merged_data2 %>% filter(sex != " unknown")


```

```{r}
# Assign labels to variables for better presentation in the table
label(merged_data2_filtered$sex) <- "Sex"
label(merged_data2_filtered$icu_status) <- "ICU Status"
label(merged_data2_filtered$age) <- "Age"
label(merged_data2_filtered$ventilator.free_days ) <- "Ventilator free Days"
label(merged_data2_filtered$hospital.free_days_post_45_day_followup) <- " hospital free days post 45 day followup"
# Convert variables to numeric if they are not already
merged_data2_filtered$age <- as.numeric(merged_data2_filtered$age)
merged_data2_filtered$ventilator.free_days <- as.numeric(merged_data2_filtered$ventilator.free_days)
merged_data2_filtered$hospital.free_days_post_45_day_followup <- as.numeric(merged_data2_filtered$hospital.free_days_post_45_day_followup)

merged_data2_filtered <- merged_data2_filtered %>%
  filter(!is.na(age) & !is.na(ventilator.free_days) & !is.na(hospital.free_days_post_45_day_followup))


# Define caption and footnote
caption <- "Table 1: Categoritcal Covariates and Continues Covariates Stratified by Status"
footnote <- "the result inculdes n (%) for categorical variables and mean (sd) or median [IQR] for continuous variables."


# Create the descriptive table stratified by 'status'
IQR_format <- c(
  "Median [IQR]" = "MEDIAN [IQR]"
)

table1(~ sex + icu_status + age + ventilator.free_days + hospital.free_days_post_45_day_followup | disease_status, 
       data=merged_data2_filtered, overall=c(left="Total"), caption=caption, footnote=footnote,render.continuous = IQR_format)
```
distribution
```{r}

library(ggplot2)
ggplot(merged_data2_filtered,aes(x = age)) + 
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7)
```


```{r}
library(ggplot2)
ggplot(merged_data2_filtered,aes(x = ventilator.free_days)) + 
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7)
```

#Distribution of hospital free days post 45 day followup

```{r}
library(ggplot2)
ggplot(merged_data2_filtered,aes(x = hospital.free_days_post_45_day_followup)) + 
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7)
```

#another plot

```{r}
# Define overall plot including the data set name and the variables you want to look at within that data frame
my_rugplot<-ggplot(merged_data,aes(x = as.numeric(age),y = ventilator.free_days,color = disease_status)) + # The + is how we communicate to ggplot2 that we aren't done adding elements to the plot
  # Add loess curve with shaded error 
  geom_smooth(method = loess) +
   labs(x = 'Age(year)',y = 'ventilator Free days',color = 'Disease Status(covid-19)', title = 'Age(year) vs. Ventilator Free Days' ) +
  # Add rug
  geom_rug()


my_rugplot

ggsave("my_rugplot.pdf", plot = my_rugplot, width = 8, height = 6)

```


```{r}
citation("tidyverse")
```


```{r}
citation
```








