---
title: "Submission 2"
output:
  pdf_document: default
  html_document: default
date: "2024-08-06"
---
```{r}
setwd("/Users/yanchuwu/Downloads") # set wd

#read csv
genes <- read.csv("QBS103_GSE157103_genes.csv",row.names = 1 ) # row name are coming from 1st colum of csv
new_genes<- t(genes)
#transform the rows to column

setwd("/Users/yanchuwu/Downloads") # set wd

#read csv

series_matrix <- read.csv("QBS103_GSE157103_series_matrix.csv")



# merge data together in to one data set file
merged_data<- merge(new_genes, series_matrix, by.x = "row.names", by.y = "participant_id", all = TRUE)
merged_data$age <- as.numeric(merged_data$age)

```


Build a function to create the plots you made for Presentation 1, incorporating any feedback you received on your submission. Your functions should take the following input: (1) the name of the data frame, (2) a list of 1 or more gene names, (3) 1 continuous covariate, and (4) two categorical covariates (10 pts)
Select 2 additional genes (for a total of 3 genes) to look at and implement a loop to generate your figures using the function you created (10 pts)

```{r}
library(ggplot2)

library(tidyverse)
library(ggpubr)

submission2_plot <- function(data, gene_name , continuous_covariate , categorical_covariates1, categorical_covariates2){
  
    
  
  # Histogram
    histogram <- ggplot(data, aes_string(x = gene_name)) + 
    geom_histogram(binwidth = 0.1, fill = "blue", alpha = 0.7) +
    labs( x = gene_name, y = "Frequency")
  
    
    #scatterplot
    data[[gene_name]] <- as.numeric(data[[gene_name]])
    
    scatter_plot <- ggplot(data, aes_string(x = continuous_covariate, y = gene_name, color = continuous_covariate)) + 
    geom_point() +
    labs(title = "gene scatter plot",
         x = continuous_covariate,
         y = gene_name)
    
    
    
    box_plot <- ggplot(data,aes_string(x = categorical_covariates1 ,y = gene_name, fill = categorical_covariates2)) +
     # Add box plot
     geom_boxplot() +
    # Change labels
     labs(x = categorical_covariates1,y = gene_name,color = categorical_covariates2 ) 

    plot(histogram)
    plot(scatter_plot)
    plot(box_plot)
    
  }

submission2_plot(data = merged_data, gene_name = 'A1BG', continuous_covariate = 'age', categorical_covariates1 = 'sex', categorical_covariates2 = 'icu_status')


```

```{r}

genes <- c('A1BG', 'A2M', 'AAAS') 

for (gene in genes) {
  submission2_plot(data = merged_data, gene_name = gene, continuous_covariate = 'age', categorical_covariates1 = 'sex', categorical_covariates2 = 'icu_status')
}

```

